---
title: "DOC Paper Q Supplement"
author: "Alex Cebulski"
format: html
date: "`r Sys.Date()`"
self-contained: true
---
  
```{r}
#| include: false

library(tidyverse)
library(sf)
library(ggforce)
library(tmap)
```


```{r}
#| include: false
#| warning: false

wsc_bsns <- st_read("data/gis/wsc_gauge_basins/hnbp_2022/hnbp_2022_wsc_basins_w_dat_avail.shp") |> 
  st_drop_geometry() 

nobGaugeArea <- wsc_bsns |> 
  select(ID = StationNum, area_km2 = Area_km2) |> 
  distinct()

nobGauge <- read.csv("data/gauged/wsc/wsc_unregulated_1981_2010_flows_wide_2022.csv") |> 
  left_join(nobGaugeArea) |> 
  select(ID,
         area = area_km2, 
         Jan:ann_mm)

nobMod <- read.csv("data/modelled/wsc/dcwbm_raw_from_amrit_wsc_unreg_1981_2010.csv") |> 
  left_join(nobGaugeArea) |> 
  select(ID,
         area = area_km2, 
         Jan:ann_mm)

#south of border

sobGauge <- read.csv("data/gauged/usgs/usgs_1981_2010_monthly_flows_wide_2022.csv") 

sobGgeArea <- sobGauge |> select(ID, area)

sobMod <- read.csv("data/modelled/usgs/dcwbm_raw_from_amrit_usgs_unreg_1981_2010.csv") |> 
  dplyr::rename(ID = GAGE_ID)  |> 
  left_join(sobGgeArea) |> 
  select(ID, area, Jan:Dec, ann_mm)

sobMobLong <- sobMod |> 
  pivot_longer(Jan:Dec, names_to = "Month", values_to = "mod_q") |> 
  mutate(ID = paste0(ID, Month))  |> 
  select(ID, mod_q)

sobGaugeLong <- sobGauge |> 
  pivot_longer(Jan:Dec, names_to = "Month", values_to = "gauge_q") |> 
  mutate(ID = paste0(ID, Month))  |> 
  select(ID, gauge_q)

# create dfs for full domain

mod <- rbind(nobMod, sobMod) 

mod_long <- mod |> 
  pivot_longer(Jan:Dec, names_to = "Month", values_to = "mod_q") |> 
  select(ID, Month, mod_q) 

gauge <- rbind(nobGauge, sobGauge) 

gauge_long <- gauge |>  
  pivot_longer(Jan:Dec, names_to = "Month", values_to = "gauge_q") |> 
  select(ID, Month, gauge_q)

all_monthly <- inner_join(mod_long, gauge_long)

all_annuals <- inner_join(mod, gauge, by = "ID") |> 
  select(ID, area = area.x, ann_mm_mod = ann_mm.x, ann_mm_gauge = ann_mm.y)

# basin meta

wsc_meta <- read.csv('data/gauged/wsc/wsc_unregulated_1981_2010_metadata.csv')
usgs_meta <- readRDS('data/gauged/usgs/usgs_1981_2010_site_meta.rds')
stn_names <- rbind(
  wsc_meta |> select(ID = STATION_NUMBER, name = STATION_NAME),
  usgs_meta |> select(ID = site_no, name = station_nm)
)

# amrit cal set 

cal_ids <- read.csv('data/modelled/raw from amrit/calibration_set.csv')$ID |> toupper()

# mapping

npctr <- read_sf('data/gis/npctr_basins/npctr_basins_valid_no_fr_clm.shp')

sf <- read_sf('data/gis/wsc_gauge_basins/hnbp_2022/hnbp_2022_wsc_all_basins.shp') 

usgs_basins <- read_sf('data/gis/usgs_gauge_basins/usgs_unregulated_basins_1981-2010_dataRetrieval_2022_update.shp')

bg <- rbind(raster::getData('GADM', country = c('CAN'), level = 0),
            raster::getData('GADM', country = c('USA'), level = 0)) %>%
  st_as_sf() %>%
  st_transform(st_crs(sf))
```

## Overview

Below are plots to be included in the DOC flux paper which uses the 1981-2010 Q norms from Amrit's regime based calibration on 38 basins in British Columbia. The plots attempt to show the quality of modelled Q through the comparison against observed values. Additional basins were included to assess uncalibrated basins and to show performance in the USA. 155 unregulated gauges were used for the first analysis. A second analysis on just the calibration set of 38 basins is shown below.

## Data Sources

Modelled Data

- 1981-2010 monthly runoff normals
- DCWBM with Amrit's calibration set

Gauged Data

- WSC: tidyhydat r package
- USGS: dataRetreival

Basin Delineations

- WSC: https://collaboration.cmc.ec.gc.ca/cmc/hydrometrics/www/HydrometricNetworkBasinPolygons/
- USGS: 
  - https://water.usgs.gov/GIS/dsdl/boundaries_shapefiles_by_aggeco.zip
  - https://water.usgs.gov/GIS/metadata/usgswrd/XML/gagesII_Sept2011.xml

Map of the 155 unregulated gauges: 

```{r}
#| echo: false
#| warning: false


tmap_options(check.and.fix = TRUE)

tm_shape(bg, bbox = npctr) +
  tm_fill() +
  tm_shape(sf |> filter(StationNum %in% mod$ID)) +
  tm_polygons(col = 'blue') +
  tm_shape(usgs_basins |> filter(GAGE_ID %in% mod$ID)) +
  tm_polygons(col = 'green') +
  tm_graticules(
  ) 
```



## Annuals

```{r}
#| warning: false
#| echo: false

ggplot(all_annuals, aes(x= ann_mm_gauge, y = ann_mm_mod)) +
  geom_point(aes()) +
  labs(x='Observed Runoff (mm)', y = 'Modelled Runoff (mm)') +
  geom_abline(intercept = 0, slope = 1) +
  xlim(0, 4000) +
  ylim(0, 4000) +
  scale_colour_brewer(palette="Dark2") +
  theme_bw() 

ggsave('plots/obs_mod_annual_runoff_1981_2010.png')
```

## Annual Statistics

```{r}
#| warning: false
#| echo: false


all_annuals |> 
  mutate(diff = ann_mm_mod - ann_mm_gauge) |> 
  summarise(
    `Mean Bias` = mean(diff, na.rm = T),
    MAE = mean(abs(diff), na.rm = T),
    RMSE = sqrt(mean(diff ^ 2, na.rm = T)),
    r_pear = cor(ann_mm_gauge, ann_mm_mod, use = 'na.or.complete', method = 'pearson'),
    r_spear = cor(ann_mm_gauge, ann_mm_mod, use = 'na.or.complete', method = 'spearman')) |> 
    mutate(across(everything(), round, digits = 2)) |> 
  knitr::kable()
```

```{r}
#| warning: false
#| echo: false


ann_stats <- all_annuals |> 
  mutate(diff = ann_mm_mod - ann_mm_gauge) |> 
  mutate(abs_diff = abs(ann_mm_mod - ann_mm_gauge)) |> 
  mutate(perc_diff = (diff / ann_mm_gauge)*100) |> 
  mutate(abs_perc_dif = (abs_diff / ann_mm_gauge)*100) |> 
  left_join(stn_names)

hist(ann_stats$perc_diff, main ="", xlab = "Percent Error", xlim = c(-100, 150))
```

## Seasonal Hydrograph Plots

```{r fig.height=8, fig.width=8,echo=FALSE, message=FALSE, warning=FALSE}
all_monthly_long <- rbind(
  mod_long |> 
    mutate(type = 'Modelled') |> 
    rename(q = mod_q),
  gauge_long |> 
    mutate(type = 'Observed') |> 
    rename(q = gauge_q)
)

# 4 good 4 bad
select_ids <- c(
                '08EG012',
                '08FB004',
                '08GA075',
                '12092000',
                '08EE008',
                '08EB005',
                '08HB032',
                '12025700'
              )

select_monthly_long <- all_monthly_long |> 
   filter(ID %in% select_ids) |> 
  left_join(stn_names)
  
# alternate good, bad for good col and bad col on plot 
select_monthly_long$ID <- factor(
  select_monthly_long$ID,
  c(
    '08EG012',
    '08FB004',
    '08GA075',
    '12092000',
    '08EE008',
    '08EB005',
    '08HB032',
    '12025700'
  )
) 

select_monthly_long_arrange <- select_monthly_long |>
              mutate(name = factor(name, levels = unique(name[order(ID)])))





# all$regime <- factor(all$regime, c('1. Pluvial', '2. Pluvial Dominant Hybrid', '3. Nival Dominant Hybrid', '4. Nival-Pluvial-Glacial', '5. Nivo-Glacial', '6. Nival', '7. Icefield'))

ggplot(select_monthly_long_arrange, aes(x = factor(Month, levels = c('Oct', 'Nov', 'Dec','Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep')), 
                y = q,
                group = type)) +
  geom_line(aes(linetype = type), size = 1) +
  labs(x='Month', y = 'Runoff (mm)') +
  theme_bw() +
  scale_colour_brewer(palette="Dark2") +
  facet_wrap_paginate(~name, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE), ncol = 2, nrow = 4, page = 1) + 
  theme(legend.title = element_blank(),
        legend.position='bottom',
        legend.box="vertical", legend.margin=margin(),
        legend.background = element_blank(),
        legend.key = element_blank(),
        axis.title.x=element_blank()) 

ggsave("plots/select_hydrographs.png", units = "in", width = 7, height = 7)

mon_stat <- 
  all_monthly |> 
  mutate(diff = mod_q - gauge_q) |> 
  group_by(ID) |> 
  summarise(
    `Mean Bias` = mean(diff, na.rm = T),
    `Max Error` = diff[which.max(abs(diff))],
    MAE = mean(abs(diff), na.rm = T),
    RMSE = sqrt(mean(diff ^ 2, na.rm = T)),
    r_pear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'pearson'),
    r_spear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'spearman'),
    nse = 1 - sum((gauge_q-mod_q)^2)/sum((gauge_q-mean(gauge_q, na.rm = T))^2), # checked and this is the same as hydroGOF package and joels code
  ) |> 
    mutate(across(-ID, round, digits = 2)) |> 
  left_join(stn_names)
```

The stats for the above plots are:

```{r fig.height=8, fig.width=8,echo=FALSE, message=FALSE, warning=FALSE}
mon_stat |> 
   filter(ID %in% select_ids) |> 
  knitr::kable()

```

## Monthly Statistics

```{r}
#| warning: false
#| echo: false

mon_stat <- 
  all_monthly |> 
  mutate(diff = mod_q - gauge_q) |> 
  group_by(ID) |> 
  summarise(
    `Mean Bias` = mean(diff, na.rm = T),
    `Max Error` = diff[which.max(abs(diff))],
    MAE = mean(abs(diff), na.rm = T),
    RMSE = sqrt(mean(diff ^ 2, na.rm = T)),
    r_pear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'pearson'),
    r_spear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'spearman'),
    nse = 1 - sum((gauge_q-mod_q)^2)/sum((gauge_q-mean(gauge_q, na.rm = T))^2), # checked and this is the same as hydroGOF package and joels code
  ) |> 
    mutate(across(-ID, round, digits = 2)) |> 
  left_join(stn_names) 

write.csv(mon_stat, 'plots/obs_mod_monthly_runoff_1981_2010_error_table.csv', row.names = F)
  
mon_stat |> knitr::kable()
```

# Calibration Set Only

38 basins were used in the 2020 regime based calibration which were all located within BC.

```{r}
#| echo: false
#| warning: false


all_annuals <- all_annuals |> filter(ID %in% cal_ids)

all_monthly <- all_monthly |> filter(ID %in% cal_ids)

sf <- sf |> filter(StationNum %in% cal_ids)

tmap_options(check.and.fix = TRUE)

tm_shape(bg, bbox = npctr) +
  tm_fill() +
  tm_shape(sf) +
  tm_polygons(col = 'blue') +
  tm_graticules(
  ) 
```


## Annuals

```{r}
#| warning: false
#| echo: false

ggplot(all_annuals, aes(x= ann_mm_gauge, y = ann_mm_mod)) +
  geom_point(aes()) +
  labs(x='Observed Runoff (mm)', y = 'Modelled Runoff (mm)') +
  geom_abline(intercept = 0, slope = 1) +
  xlim(0, 4000) +
  ylim(0, 4000) +
  scale_colour_brewer(palette="Dark2") +
  theme_bw() 

ggsave('plots/obs_mod_annual_runoff_1981_2010_cal_only.png')
```

## Annual Statistics

```{r}
#| warning: false
#| echo: false


all_annuals |> 
  mutate(diff = ann_mm_mod - ann_mm_gauge) |> 
  summarise(
    `Mean Bias` = mean(diff, na.rm = T),
    MAE = mean(abs(diff), na.rm = T),
    RMSE = sqrt(mean(diff ^ 2, na.rm = T)),
    r_pear = cor(ann_mm_gauge, ann_mm_mod, use = 'na.or.complete', method = 'pearson'),
    r_spear = cor(ann_mm_gauge, ann_mm_mod, use = 'na.or.complete', method = 'spearman')) |> 
    mutate(across(everything(), round, digits = 2)) |> 
  knitr::kable()
```

```{r}
#| warning: false
#| echo: false


ann_stats <- all_annuals |> 
  mutate(diff = ann_mm_mod - ann_mm_gauge) |> 
  mutate(abs_diff = abs(ann_mm_mod - ann_mm_gauge)) |> 
  mutate(perc_diff = (diff / ann_mm_gauge)*100) |> 
  mutate(abs_perc_dif = (abs_diff / ann_mm_gauge)*100) |> 
  left_join(stn_names)

hist(ann_stats$perc_diff, main ="", xlab = "Percent Error", xlim = c(-100, 150))
```

## Seasonal Hydrograph Plots

```{r fig.height=8, fig.width=8,echo=FALSE, message=FALSE, warning=FALSE}
all_monthly_long <- rbind(
  mod_long |> 
    mutate(type = 'Modelled') |> 
    rename(q = mod_q),
  gauge_long |> 
    mutate(type = 'Observed') |> 
    rename(q = gauge_q)
)

# 4 good 4 bad
select_ids <- c(
                '08EG012',
                '08FB004',
                '08GA075',
                '12092000',
                '08EE008',
                '08EB005',
                '08HB032',
                '12025700'
              )

select_monthly_long <- all_monthly_long |> 
   filter(ID %in% select_ids) |> 
  left_join(stn_names)
  
# alternate good, bad for good col and bad col on plot 
select_monthly_long$ID <- factor(
  select_monthly_long$ID,
  c(
    '08EG012',
    '08FB004',
    '08GA075',
    '12092000',
    '08EE008',
    '08EB005',
    '08HB032',
    '12025700'
  )
) 

select_monthly_long_arrange <- select_monthly_long |>
              mutate(name = factor(name, levels = unique(name[order(ID)])))





# all$regime <- factor(all$regime, c('1. Pluvial', '2. Pluvial Dominant Hybrid', '3. Nival Dominant Hybrid', '4. Nival-Pluvial-Glacial', '5. Nivo-Glacial', '6. Nival', '7. Icefield'))

ggplot(select_monthly_long_arrange, aes(x = factor(Month, levels = c('Oct', 'Nov', 'Dec','Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep')), 
                y = q,
                group = type)) +
  geom_line(aes(linetype = type), size = 1) +
  labs(x='Month', y = 'Runoff (mm)') +
  theme_bw() +
  scale_colour_brewer(palette="Dark2") +
  facet_wrap_paginate(~name, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE), ncol = 2, nrow = 4, page = 1) + 
  theme(legend.title = element_blank(),
        legend.position='bottom',
        legend.box="vertical", legend.margin=margin(),
        legend.background = element_blank(),
        legend.key = element_blank(),
        axis.title.x=element_blank()) 

ggsave("plots/select_hydrographs_cal_only.png", units = "in", width = 7, height = 7)

mon_stat <- 
  all_monthly |> 
  mutate(diff = mod_q - gauge_q) |> 
  group_by(ID) |> 
  summarise(
    `Mean Bias` = mean(diff, na.rm = T),
    `Max Error` = diff[which.max(abs(diff))],
    MAE = mean(abs(diff), na.rm = T),
    RMSE = sqrt(mean(diff ^ 2, na.rm = T)),
    r_pear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'pearson'),
    r_spear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'spearman'),
    nse = 1 - sum((gauge_q-mod_q)^2)/sum((gauge_q-mean(gauge_q, na.rm = T))^2), # checked and this is the same as hydroGOF package and joels code
  ) |> 
    mutate(across(-ID, round, digits = 2)) |> 
  left_join(stn_names)
```

The stats for the above plots are:

```{r fig.height=8, fig.width=8,echo=FALSE, message=FALSE, warning=FALSE}
mon_stat |> 
   filter(ID %in% select_ids) |> 
  knitr::kable()

```

## Monthly Statistics

```{r}
#| warning: false
#| echo: false

mon_stat <- 
  all_monthly |> 
  mutate(diff = mod_q - gauge_q) |> 
  group_by(ID) |> 
  summarise(
    `Mean Bias` = mean(diff, na.rm = T),
    `Max Error` = diff[which.max(abs(diff))],
    MAE = mean(abs(diff), na.rm = T),
    RMSE = sqrt(mean(diff ^ 2, na.rm = T)),
    r_pear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'pearson'),
    r_spear = cor(gauge_q, mod_q, use = 'na.or.complete', method = 'spearman'),
    nse = 1 - sum((gauge_q-mod_q)^2)/sum((gauge_q-mean(gauge_q, na.rm = T))^2), # checked and this is the same as hydroGOF package and joels code
  ) |> 
    mutate(across(-ID, round, digits = 2)) |> 
  left_join(stn_names) 

write.csv(mon_stat, 'plots/obs_mod_monthly_runoff_1981_2010_error_table_cal_only.csv', row.names = F)
  
mon_stat |> knitr::kable()
```